Stage-Splitting a Modal Language

### 2014-10-30 paper/splitting.tex
heavy edits to 4.0-4.2

@@

%As with the dynamics, our stage-splitting algorithm for expressions in \lang\
%takes the form of two mutually-dependent judgments, $\splitonesym$ and
%$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
%\bbtwo$.  Before diving into the full generality,

The goal of splitting is to take a multistage \lang\ term,
which may feature interleaved stage \bbone\ and stage \bbtwo\ code,
and untangle it to produce two separate terms:
one with all of the stage \bbone\ CONTENT, and one with all of
the stage \bbtwo\ CONTENT.

Splitting emits terms in an unstaged language, \langmono, that is 
untyped and contains no staging features, but otherwise has the same term constructors as \lang. 
\langmono\ has a evaluation judgment $\redsym$, which works in the expected way to produce unstaged values.

Since splitting stage \bbtwo\ terms from \lang\ is a simpler operation than splitting stage 1 terms,
we first describe the mechanics of splitting stage \bbtwo, before moving to the more complex
case of stage \bbone\ term splitting.

%The splitting algorithm comes in several forms, depending on the stage of the input and whether it is a value.
%The stage \bbtwo\ versions will be addressed first,
%before the more complicated stage \bbone\ versions.

@@

% We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.

Splitting a stage-2 term $\coltwo e A$ produces a term
$p$ (called the {\em precomputation}) and a term $l.r$ (called the {\em resumer})
open on a single variable, such that $e$ is equivalent to $\letin l p r$.
Intuitively, the responsibility of $p$ is to produce an intermediate data structure, 
and the responsibility of $r$ is to consume that data structure and produce the same final value as $e$.

 For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a resumer
\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to
the precomputation must agree with the original term. This process of hoisting
all the stage \bbone\ subterms out of a term is called \emph{precomputation lifting}.

Splitting is implemented by the judgment $\splittwosym$,
the rules for which are given in \cref{fig:termSplitTwo}.  
It operates inductively on the structure of the term.  
As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
from contained stage \bbone\ code into a single expression $p$, 
leaving a resumer $l.r$ to interpret the results of that precomputation.
Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
that is, $p$ is a unit value and $l$ is never used in $r$.
At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
the precomputations of the constituent parts, and then unbundling them with a pattern,
with the resumer precisely resembling the original expression.
The $\prev$ rule depends on splitting stage \bbone\ terms, which is covered later.

@@

Observe that, when applied to stage \bbtwo\ values, 
splitting always produces an effectively trivial precomputation:
either \texttt{()}, or nested tuples thereof.
Intuitively, this makes sense---there is no way to embed stage \bbone\ code
in the internals of a stage \bbtwo\ value, 
thus they can have no stage \bbone\ content.
Indeed it would be almost nonsensical to talk of an algorithm for splitting stage \bbtwo\ values,
since the precondition, that the input is a stage \bbtwo\ value, is enough to imply the postcondition,
that no stage interleaving remains.

The same is not true of partial values---partial values, a restriction on stage \bbone\ terms,
{\em can} have stage \bbtwo\ content.
Thus, it is meaningful to define some notion of splitting for them.
This is useful to study, because the definition will later be used as a contract 
with which general stage \bbone\ term splitting must comply.

Recall that a partial value is externally a stage \bbone\ thing; 
it has a type $@~\bbone$, and along with a future context, it is the result of reducing a stage \bbone\ term.
But internally, a partial value might contain stage \bbtwo\ subexpressions.
There are two places where such embedding may occur: 
$\next$ blocks, which must contain a variable reference to some future context; 
and lambda values, which may contain a $\next$ in their body.
The goal of splitting for stage \bbone\ partial values is therefore 
to separate the contents of $\next$ blocks from the rest of the value, 
while maintaining some structure.

Splitting a partial value $v$ creates two unstaged values $v_1$ and $v_2$,
where $v_w$ contains only the information required for stage $w$ evaluation of
$v$. Splitting partial values is implemented by the judgment $v \vsplito
[v_1,v_2]$ defined in \ref{fig:valSplit}. 
At product, sum, and recursive types, splitting acts recursively, gathering all
subterms' stage $w$ content into $v_w$. At base types, splitting assigns $v$
entirely to stage \bbone\ (and \texttt{()} to stage \bbtwo); on the other hand,
at $\fut$ types, splitting assigns $v$ (which must have the form $\next~\hat y$)
entirely to stage \bbtwo\ (and \texttt{()} to stage \bbone). This process of
exposing only the relevant structure of a term is called \emph{masking}.

@@

>>>>

%\langmono\ has a evaluation judgment $\redsym$, which works in the expected way
%to produce unstaged values.

@@

%As with the dynamics, our stage-splitting algorithm for expressions in \lang\
%takes the form of two mutually-dependent judgments, $\splitonesym$ and
%$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
%\bbtwo$.

Although terms in \lang\ clearly distinguish their stage \bbone\ and stage
\bbtwo\ components, these components may be interleaved in complex ways. The
goal of \emph{stage splitting} is to produce from a term in \lang\ a pair of
terms in an unstaged language \langmono, which respectively perform all the
stage \bbone\ and stage \bbtwo\ subcomputations of the original term.

We begin by describing how to split stage \bbtwo\ terms.

%Since splitting stage \bbtwo\ terms from \lang\ is a simpler operation than
%splitting stage 1 terms, we first describe the mechanics of splitting stage
%\bbtwo, before moving to the more complex case of stage \bbone\ term splitting.

@@

Splitting a stage-\bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
term $p$ and a \emph{resumer} function $l.r$, such that $e$ is equivalent to
$\letin l p r$. This is implemented by the judgment $\splittwosym$,
(\cref{fig:termSplitTwo}), which operates inductively on the structure of the
term.

For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a resumer
\verb|(L1,L2) => L1 < L2-5|, with the property that the resumer applied to the
precomputation must agree with the original term. This process of hoisting stage
\bbone\ subterms out of $e$ is called \emph{precomputation lifting}, and
intuitively, amounts to computing a data structure which is consumed by $r$.

Since base types and variables have no stage \bbone\ content, their
precomputations are \texttt{()}, and are not used in the resumer. At every other
expression except $\prev$, splitting bundles the precomputations of the
constituent parts, and then unbundles them with a pattern, with the resumer
precisely resembling the original expression. The $\prev$ rule depends on
splitting stage \bbone\ terms, which is covered later.

%As mentioned before, the only responsibility of $\splittwosym$ is to lift
%out the precomputation from contained stage \bbone\ code into a single
%expression $p$, leaving a resumer $l.r$ to interpret the results of that
%precomputation.

@@

As described above, splitting a stage \bbtwo\ value always produces a trivial
precomputation---either \texttt{()}, or nested tuples thereof. Intuitively,
this is because stage-\bbtwo\ values contain no stage \bbone\ computations.

The same is not true of partial values (at stage \bbone), which \emph{can}
contain stage \bbtwo\ computations, namely, inside $\next$ blocks, which must
contain a variable reference to some future context, and inside lambdas, which
may contain a $\next$ in their body.

Thus, to split partial values, we must separate the contents of $\next$ blocks
from the rest of the term, while maintaining the term's original structure. This
is implemented by the judgment $v \vsplito [v_1,v_2]$ (\ref{fig:valSplit}).
At product, sum, and recursive types, splitting acts recursively, gathering all
subterms' stage $w$ content into $v_w$.
At base types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()}
to stage \bbtwo);
on the other hand, at $\fut$ types, splitting assigns $v$ (which must have the
form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
\bbone).
This process of exposing only the relevant structure of a term at each stage is
called \emph{masking}.

@@

%Indeed it would be almost nonsensical to talk of an algorithm for splitting
%stage \bbtwo\ values, since the precondition, that the input is a stage \bbtwo\
%value, is enough to imply the postcondition, that no stage interleaving remains

%Thus, it is meaningful to define some notion of splitting for them.  This is
%useful to study, because the definition will later be used as a contract with
%which general stage \bbone\ term splitting must comply.

%Recall that a partial value is externally a stage \bbone\ thing; it has a type
%$@~\bbone$, and along with a future context, it is the result of reducing a
%stage \bbone\ term. But internally, a partial value might contain stage \bbtwo\
%subexpressions.
